---
title: Explaining Patreon's Subscription Changes
author: Brian Richards
date: '2017-12-08'
slug: explaining-patreon-s-subscription-changes
categories: []
tags: []
draft: yes
comments: false
share: false
---

On December 8th, 2017, I noticed that my Twitter feed was blowing up about 
the [new changes coming to Patreon](https://blog.patreon.com/updating-patreons-fee-structure/). 
The big change that has people nervous is that starting in late December, 
Patreon patrons would have to start paying for credit card service fees. This 
is a change over the previous system where all fees were taken out of the 
amount the creator received directly. 
While overall a switch to charging the patrons the transaction fees doesn't 
seem like a huge change, the deeper change of interest is the disaggregation 
of the transaction fees.

The trait that made Patreon different from other services was that you could 
aggregate the transaction fees. With any credit card transaction, there is 
typically a fee of some percentage of the charge plus a standard cost. For 
Patreon, this was a 1.9% + \$0.30 fee that was subtracted from the money sent 
to the creators. This was done once a month for all of your pledges, so if you 
pledge to a number of different creators the transaction fee would be spread 
across them. Under the new system, Patreon is going to charge a 2.9% + 
\$0.35 on top of each pledge made. This means that a person that pledges \$1 
to 10 different creators is going to be paying a \$3.79 transaction fee where 
previously that fee was \$0.64. 

There are a few major concerns in the community regarding this. Some of the 
main ones being that many creators make their money from small pledges and 
the additional service fee might make those patrons leave. Some patrons may 
only have so much they can spend on Patreon per month, so those patrons may 
drop some of their small pledges or leave the platform entirely. 

When I first read about this, I didn't realize that they were going to 
disaggregate the pledge payments. But then I saw this Twitter post from 
\@FoldableHuman that included a picture from the updated Patreon blog post 
about the changes: 

```{r, echo = FALSE}
blogdown::shortcode('tweet', '939007425188847616')
```

That image was my first hint that Patreon was disaggregating the transaction 
fee in addition to having the patrons pay for it. I also saw a nice figure 
by \@DarwinAwdWinner showing how much of what the patron is paying for a 
\$1 pledge goes to the transaction fee, the Patreon cut, and the creator cut:

```{r, echo = FALSE}
blogdown::shortcode('tweet', '938837508552400896')
```

These two figure got me thinking. What are these changes doing to the power of 
each dollar I spend on the service?

*Note that in the following section I will be talking about the percentage of 
the total money paid by the patron that is given to the creator*

## Leveraging Patron's Money

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
```

After reading about the changes in Patreon's subscription and fee structure, 
the following question came to me: 

> Is the power of my money spent being reduced by the loss of fee aggregation?

To figure that out, let's do a calculation. For simplicity, I am going to 
only use the current Patreon fee of 5% and the current Stripe payment processing 
fee of 1.9% + $0.30 per transaction as that is most similar to the new proposed 
fee structure.

Under the old structure, Patreon would aggregate the fees for pledge 
transactions. What this means is that at the first of each month, they would 
charge your credit card only once for all of your pledges, subtract off the 
transaction fee, and then divide the remaining amount among the creators you 
pledged to (after taking their 5% cut). 

Under the new system, Patreon removes the aggregation of fees and will charge 
your credit card for each pledge. This means that each pledge will feel the 
full brunt of the new transaction fee (2.9% + $0.35). 

To understand how this affects the patron, we will look at a simplified 
example.

### Old System

Let us assume that the patron gives the same amount to each 
creator. For the old payment system, we could model it something like this. 

```{r, echo = FALSE}
old_payment <- function(pledge_amount, number_supported) {
  stripe_fee <- pledge_amount * number_supported * 0.019 + 0.3
  patreon_cut <- pledge_amount * 0.05
  
  breakdown <- tibble(old_stripe_creator_cut = pledge_amount - stripe_fee / number_supported - patreon_cut, 
                    old_stripe_patron_paid = pledge_amount, 
                    old_stripe_creator_percent = (pledge_amount - stripe_fee / number_supported - patreon_cut) / pledge_amount * 100)  
}
```

Here we say that the user pledges some amount to n creators. We can break down 
the amount the user pays and the amount the creator makes as follows: 

* The cut the creator made (after fees and Patreon cut) 
* The amount the patron paid 
* The percentage of money the creator gets from what the patron paid

This last breakdown is the most important for me as it gives a hint to how much 
of the money the patron spends is actually going to the creator. Now, let's 
calculate the breakdown over a range of pledges from \$1 to \$20 and a range 
of creators supported from 1 to 20.

```{r, echo = FALSE}
old_payment_calculation <- tibble(pledge = c(1:20)) %>% 
  mutate(list(tibble(creators = c(1:20)))) %>% 
  unnest() %>% 
  group_by(pledge, creators) %>% 
  mutate(list(old_payment(pledge, creators))) %>% 
  unnest() %>% 
  ungroup()
```

We can now plot the percentage of the patrons money that the creators will take 
home. First we will look at the percentage taken home as a function of the 
amount pledged. This plot is color coded for the number of creators supported.

```{r, echo = FALSE}
old_payment_calculation %>% 
  filter(creators %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = pledge, 
             y = old_stripe_creator_percent, 
             color = as.factor(creators))) + 
  geom_line(size = 1) +
  theme_bw() + 
  labs(x = "Amount Pledged (USD)", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Number of\nCreators\nSupported") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Here we see that the percentage that the creator takes home compared to what 
the patron actually paid increases with increases in the amount pledged. 
Additionally, as the number of creators supported increases, the percentage 
of low pledge amounts taken home drastically increases. 

We can compare that to a plot of the percentage taken home by the creator as 
a function of the number of creators supported with color showing the 
amount of money pledged.

```{r, echo = FALSE}
old_payment_calculation %>% 
  filter(pledge %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = creators, 
             y = old_stripe_creator_percent, 
             color = as.factor(pledge))) + 
  geom_line(size = 1) +
  theme_bw() + 
  labs(x = "Number of Creators Supported", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Amount Pledged\n(USD)") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Again, we see a similar trend of increase percentages for more creators being 
supported.

### New System

Now, we want to calculate the same break down, but for the new system of 
subscriptions and transactions. A key note here, the amount pledged by the 
patron and the amount actually paid by the patron will be different. Under the 
new system, Patreon will add a 2.9% + $0.35 service fee onto every pledge and 
the pledges will be transacted separately.

```{r, echo = FALSE}
new_payment <- function(pledge_amount, number_supported) {
  transaction_fee <- pledge_amount * 0.029 + 0.35
  patreon_cut <- pledge_amount * 0.05
  
  breakdown <- tibble(new_creator_cut = pledge_amount - patreon_cut, 
                    new_patron_paid = pledge_amount + transaction_fee, 
                    new_creator_percent = (pledge_amount - patreon_cut) / new_patron_paid * 100)  
}
```

As with the old system of payment, let's calculate the breakdown over a range 
of pledges from \$1 to \$20 and a range of creators supported from 1 to 20.

```{r, echo = FALSE}
new_payment_calculation <- tibble(pledge = c(1:20)) %>% 
  mutate(list(tibble(creators = c(1:20)))) %>% 
  unnest() %>% 
  group_by(pledge, creators) %>% 
  mutate(list(new_payment(pledge, creators))) %>% 
  unnest() %>% 
  ungroup()
```

We can now recreate the figures above using the new calculation.
First we will look at the percentage taken home as a function of the 
amount pledged. This plot is color coded for the number of creators supported.

```{r, echo = FALSE}
new_payment_calculation %>% 
  filter(creators %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = pledge, 
             y = new_creator_percent, 
             color = as.factor(creators))) + 
  geom_line(size = 1, position = "jitter") +
  theme_bw() + 
  labs(x = "Amount Pledged (USD)", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Number of\nCreators\nSupported") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Here we see that the percentage that the creator takes home compared to what 
the patron actually paid increases with increases in the amount pledged, just 
as in the old case. However, the number of creators supported is no longer 
coupled to the service feed. This means that there is no link between the 
percentage of amount paid by the patron that the creator takes home and the 
number of creators supported. This is seen here by all of the different color 
lines for the number of creators supported overlapping. The lines themselves 
were jittered so that the overlap would be clear.

This becomes even clearer when we look at the percentage of amount paid by the 
patron as a function of the number of creators supported.

```{r, echo = FALSE}
new_payment_calculation %>% 
  filter(pledge %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = creators, 
             y = new_creator_percent, 
             color = as.factor(pledge))) + 
  geom_line(size = 1) +
  theme_bw() + 
  labs(x = "Number of Creators Supported", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Amount Pledged\n(USD)") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Now we have a flat line for each level. Disaggregating the user payments means 
that increasing the number of creators supported makes no change in the percent 
that creators take home from what the patrons pay.

### Comparison

Let's put these side by side for easier comparison.

```{r, echo = FALSE}
comparison <- old_payment_calculation %>% 
  left_join(new_payment_calculation, 
            by = c("pledge", "creators")) %>% 
  gather(system, creator_percent, old_stripe_creator_percent, new_creator_percent) %>% 
  mutate(system = factor(system, levels = c("old_stripe_creator_percent", "new_creator_percent"), labels = c("Old", "New")))

```

First, let's take a look at the precentage of amount paid to the creator as 
a function of amount pledged for both systems.

```{r, echo = FALSE}
comparison %>% 
  filter(creators %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = pledge, 
             y = creator_percent, 
             color = as.factor(creators))) + 
  geom_line(size = 1) +
  geom_point() +
  facet_grid(. ~ system) +
  theme_bw() + 
  labs(x = "Amount Pledged (USD)", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Number of\nCreators\nSupported") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Here it is clear that no matter how many creators you support under the new 
system, you still get a percentage very close to only supporting a single 
creator under the old system.

```{r, echo = FALSE}
comparison %>% 
  filter(pledge %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = creators, 
             y = creator_percent, 
             color = as.factor(pledge))) + 
  geom_line(size = 1) +
  geom_point() +
  facet_grid(. ~ system) +
  theme_bw() + 
  labs(x = "Number of Creators Supported", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Amount Pledged\n(USD)") +
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

The story is similar here when looking as a function of the number of creators 
supported. Your money as a patron goes further for the creators the more 
creators you support under the old system. Under the new system, more of your 
money is going into fees if you support multiple creators with small pledges. 

Let me say that again. **Under the new system, more of the patron's money 
goes into fees if you support multiple creators with small pledges.**

Of course, my scenario above is a generalization. There are likely not patrons 
pledging the same value to all creators. Some are probably pledging more to 
some creators than others. However, any aggregation of the transaction fees 
is a benefit to the creators and the patrons both.

## A Note on Patreon's Reasoning

In reading through their blog post, I understand why Patreon made the 
decision that they did. Their decision stems from a desire to allow creators to 
require patrons to pay first before being able to access their rewards. But 
this apparently led to a problem of perceived double booking, where patrons 
got charged immediately and then were charged shortly after when the first of 
the month rolled around. To fix this problem, Patreon decided to switch to a 
new subscription system that removed the possibility to aggregate fees because 
every pledge would now be a separate subscription with its own transaction that 
occured monthly depending on when you initially pledged. So if you pledged to 
four different creators, you now have four different transactions instead of 
all of the transactions being aggregated into one.

In Patreon's own words: 

> "...we need to move our payments system to treat your pledges like *any other 
> subscription* service. In other words, we need a system that charges patrons at 
> the time of their initial pledge, and on the anniversary of their pledge each 
> month thereafter."

For me, the part where they have gone wrong on this is that Patreon isn't like 
any other subscription service. It is an opportunity for creators to get direct 
help from their fans and make use of large fanbases that might not be able to 
pay large amounts of money. By aggregating the transaction fees, they are able 
to help get more money into the pockets of the creators.

Patreon claims that they are doing this with intention of getting the most 
money for creators. I think that is a goal that everyone can agree on. But for 
me, my goal is to get as much money as I can directly to the creators. Small 
pledges are hard to work because of the flat fee attached to transaction fees. 
By aggregating the transactions, Patreon helped make these work for creators. 
Under the new system, they aren't encouraging me to seek out new creators to 
back. Doing so only hits me with more fees and doesn't help the rest of the 
creators that I support. 

I plan on continuing to support the creators that I already 
follow and I hope that these changes won't hurt them too much. And who knows, 
this sounds like a great opportunity for a competitor to make a name for 
itself in a niche that Patreon has had a pretty strong hold of for several 
years.
