---
title: Explaining Patreon's Subscription Changes
author: Brian Richards
date: '2017-12-08'
slug: explaining-patreon-s-subscription-changes
categories: []
tags: []
draft: yes
---

```{r, message = FALSE}
library(tidyverse)
```

## Leveraging User's Money

After reading about the changes in Patreon's subscription and fee structure, 
the following question came to me: 

> Is the power of my money spent being reduced by the loss of fee aggregation?

To figure that out, let's do a calculation. For simplicity, I am going to 
only use the current Patreon fee of 5% and the current Stripe payment processing 
fee of 1.9% + $0.30 per transaction as that is most similar to the new proposed 
fee structure.

Under the old structure, Patreon would aggregate the fees for pledge 
transactions. What this means is that at the first of each month, they would 
charge your credit card only once for all of your pledges, subtract off the 
transaction fee, and then divide the remaining amount among the creators you 
pledged to (after taking their 5% cut). 

Under the new system, Patreon removes the aggregation of fees and will charge 
your credit card for each pledge. This means that each pledge will feel the 
full brunt of the new transaction fee (2.9% + $0.35). 

To understand how this affects the patron, we will look at a simplified 
example.

### Example: Same pledge amount

#### Old System

Let us assume that the patron gives the same amount to each 
creator. For the old payment system, we could model it something like this. 

```{r}
old_payment <- function(pledge_amount, number_supported) {
  stripe_fee <- pledge_amount * number_supported * 0.019 + 0.3
  patreon_cut <- pledge_amount * 0.05
  
  breakdown <- tibble(old_stripe_creator_cut = pledge_amount - stripe_fee / number_supported - patreon_cut, 
                    old_stripe_patron_paid = pledge_amount, 
                    old_stripe_creator_percent = (pledge_amount - stripe_fee / number_supported - patreon_cut) / pledge_amount * 100)  
}
```

Here we say that the user pledges some amount to n creators. We can break down 
the amount the user pays and the amount the creator makes as follows: 

* The cut the creator made (after fees and Patreon cut) 
* The amount the patron paid 
* The percentage of money the creator gets from what the patron paid

This last breakdown is the most important for me as it gives a hint to how much 
of the money the patron spends is actually going to the creator. Now, let's 
calculate the breakdown over a range of pledges from \$1 to \$20 and a range 
of creators supported from 1 to 20.

```{r}
old_payment_calculation <- tibble(pledge = c(1:20)) %>% 
  mutate(list(tibble(creators = c(1:20)))) %>% 
  unnest() %>% 
  group_by(pledge, creators) %>% 
  mutate(list(old_payment(pledge, creators))) %>% 
  unnest() %>% 
  ungroup()
```

We can now plot the percentage of the patrons money that the creators will take 
home. First we will look at the percentage taken home as a function of the 
amount pledged. This plot is color coded for the number of creators supported.

```{r}
old_payment_calculation %>% 
  filter(creators %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = pledge, 
             y = old_stripe_creator_percent, 
             color = as.factor(creators))) + 
  geom_line(size = 1) +
  theme_bw() + 
  labs(x = "Amount Pledged (USD)", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Number of\nCreators\nSupported") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Here we see that the percentage that the creator takes home compared to what 
the patron actually paid increases with increases in the amount pledged. 
Additionally, as the number of creators supported increases, the percentage 
of low pledge amounts taken home drastically increases. 

We can compare that to a plot of the percentage taken home by the creator as 
a function of the number of creators supported with color showing the 
amount of money pledged.

```{r}
old_payment_calculation %>% 
  filter(pledge %in% c(1, 2, 4, 10, 20)) %>% 
  ggplot(aes(x = creators, 
             y = old_stripe_creator_percent, 
             color = as.factor(pledge))) + 
  geom_line(size = 1) +
  theme_bw() + 
  labs(x = "Number of Creators Supported", 
         y = "Percentage of Amount Paid\n(Creators Take / Amount Paid) * 100", 
         color = "Amount Pledged\n(USD)") + 
  scale_color_brewer(palette = "Dark2") + 
  ylim(c(NA, 100))
```

Again, we see a similar trend of increase percentages for more creators being 
supported.






```{r}
old_payment_stripe <- function(pledge_amount) {
  stripe_fee <- sum(pledge_amount) * 0.019 + 0.3
  patreon_cut <- pledge_amount * 0.05
  
  breakdown <- tibble(pledge = pledge_amount, 
                    old_stripe_creator_cut = pledge_amount - stripe_fee / length(pledge_amount) - patreon_cut, 
                    old_stripe_patron_paid = pledge_amount, 
                    old_stripe_creator_percent = (pledge_amount - stripe_fee / length(pledge_amount) - patreon_cut) / pledge_amount)  
  return(breakdown)
}
```

```{r}
new_payment <- function(pledge_amount) {
  transaction_fee <- pledge_amount * 0.029 + 0.35
  patreon_cut <- pledge_amount * 0.05
  
  breakdown <- tibble(pledge = pledge_amount, 
                    new_creator_cut = pledge_amount - patreon_cut, 
                    new_patron_paid = pledge_amount + transaction_fee, 
                    new_creator_percent = (pledge_amount - patreon_cut) / new_patron_paid)  
  return(breakdown)
}
```
