---
title: Playing With The xray Package
author: Brian Richards
date: '2018-01-02'
slug: playing-with-the-xray-package
categories:
  - Data Science
  - R
tags:
  - Data Science
  - R
draft: yes
share: false
comments: false
---

After seeing an [announcement about it on R-Bloggers](https://blog.datascienceheroes.com/x-ray-vision-on-your-datasets/), 
I decided to test out the new `xray` package using the Titanic data set. The 
`xray` package provides a few functions for quickly getting a summary of 
anomalies and distributions of the variables in a data set. For anomalies, the 
`anomalies()` function outputs the number and percentage of NA's, zeroes, empty 
strings, and infinities while also giving some useful information about 
distinct observations and variable class. For distributions, the 
`distributions()` functions outputs boxplots and histograms of each variable. 
For numerical variables, the histograms also have a density curve overlay. 
These plots give a good starting point for further analysis.

I used the Titanic data set hosted via Kaggle to play around with the `xray` 
package. Below is a [partial reproduction of my Kaggle kernel](https://www.kaggle.com/bgrich/titanic-eda-using-xray-package) showing 
an exploratory data analysis of the Titanic data set using `xray`. For the full 
analysis, see the linked kernel.

# Titanic EDA Using `xray` Package

To get a sense for the Titanic data set, we are going to do some exploratory 
data analysis. As part of that, we will use the `xray` package to check 
for anomalies and look at distributions.

```{r setup, message = FALSE}
library(tidyverse)
library(xray)
library(knitr)
```

## Data Import

First, we need to import the training and testing data.

```{r, message = FALSE}
training <- read_csv("train.csv", progress = FALSE)
testing <- read_csv("test.csv", progress = FALSE)
```

Now, let's take a look at the data.

```{r}
training
```

We can also use `glimpse` to take a look at the variables. 

```{r}
glimpse(training)
```

We have 12 variables that we will want to look at. According to the data set 
description from Kaggle, the variables are defined in the following way: 

* PassengerId -- A numerical id for each passenger.
* Survived -- The response variable as a factor showing survival.
* Pclass -- The ticket class of the passenger used as a proxy for socio-economic 
class. This variable is a factor with levels 1 = 1st (Upper Class), 
2 = 2nd (Middle Class), and 3 = 3rd (Lower Class).
* Name -- The full name of the passenger with title. Married female passengers 
appear to have their partner's name listed with their actual name in 
parentheses.
* Sex -- A factor showing the sex of the passenger.
* Age -- The age of the passenger. Estimated ages are given as xx.#. 
* SibSp -- The number of siblings and spouses the passnger has aboard. 
* Parch -- The number of parent/children relationships that the passneger has 
aboard.
* Ticket -- Ticket number of the passenger.
* Fare -- The passenger fare. 
* Cabin -- The passenge cabin number. 
* Embarked -- The port of embarkation. This is a factor with three levels 
C = Cherbourg, Q = Queenstown, S = Southampton.

Before we begin the exploration, let's recode the factors with proper levels. 

```{r}
training <- training %>% 
  mutate(Survived = factor(Survived, 
                           levels = c(1, 0), 
                           labels = c("Yes", "No"))) %>% 
  mutate(Pclass = factor(Pclass, 
                         levels = c(3, 2, 1), 
                         labels = c("3rd", "2nd", "1st"), 
                         ordered = TRUE)) %>% 
  mutate(Sex = factor(Sex, 
                      levels = c("male", "female"))) %>% 
  mutate(Embarked = factor(Embarked, 
                           levels = c("C", "Q", "S"), 
                           labels = c("Cherbourg", "Queenstown", "Southampton")))
```

```{r}
testing <- testing %>% 
  mutate(Pclass = factor(Pclass, 
                         levels = c(3, 2, 1), 
                         labels = c("3rd", "2nd", "1st"), 
                         ordered = TRUE)) %>% 
  mutate(Sex = factor(Sex, 
                      levels = c("male", "female"))) %>% 
  mutate(Embarked = factor(Embarked, 
                           levels = c("C", "Q", "S"), 
                           labels = c("Cherbourg", "Queenstown", "Southampton")))
```

## Exploration with `xray`

### Checking Anomalies with `xray`

Now that the data sets have been re-coded with factors, let's take a look for 
any anomalies in the data set using the `xray` package. The `xray` package 
gives access to several functions for performing exploratory data analysis. 
The first one we will use is the `anomalies()` function that returns a data 
frame summarizing the variables in a provided data frame. 

The function returns summaries for each variable showing:

* The number of observations
* The number and percentage of observations that are NA
* The number and percentage of observations that are zero
* The number and percentage of observations that are blank
* The number and percentage of observations that are infinite
* The number of distinct observations
* The variable type 
* The percentage of observations that are "anomalous"

This information can give us a starting point to examine the data set and see 
what variables will be useful for training our models.

```{r xray_anomalies, results = 'asis'}
xray::anomalies(training)$variables %>% 
  knitr::kable()
```

From calculating the anomalies on the training data, we can see that most of 
the entries in the Cabin variable are NA. There are two options with this 
entry as we can either remove the Cabin variable entirely or recode all of the 
NA's as "No Cabin" or some similar value. We will investigate the other two 
variables with NA's below to see if we can turn them into something useful.

There are also three variables that show a number of zero observations. For the 
Parch and SibSp variables, this should be fine. This just means that those 
passengers are traveling without family members. For the fare, we may want to 
investigate which of the passengers have zero fare as they may be crew members 
or have some other identifying information.

One interesting point that pops out from the anomalies function is that the 
Ticket variable has only 681 distinct observations. This means that there must 
be a few repeat tickets that bears a quick investigation.

Note: As of this writing, the `xray` does not currently support ordered factors. 
This is why the `Pclass` variable has a class of "Unknown" from the anomalies() 
function. This will also cause a problem with the distributions() function so 
we will remove the ordering for that.

### Distributions with `xray`

Next, we will use the `distributions()` function from the `xray` package to 
visually inspect the variables. This function returns a series of box plots 
(for categorical variables) and histograms (for numerical variables) along 
with giving some distributions and breakdowns for the numerical variables. 

For the bar plots, the axis labeled as "Rows" is actually the number of 
observations of each variable category. For the histograms, the number of 
observations has been scaled for the associated density plot.

```{r, warning = FALSE, fig.width = 9}
training %>% 
  select(-PassengerId, -Name) %>% 
  mutate(Pclass = factor(Pclass, ordered = FALSE)) %>% 
  xray::distributions()
```

The `distributions()` function gives us a very rough visual tool for inspecting 
the variables. The function allows us to see some general trends and places 
that could use further checking. For instance, with this data set, it is clear 
that the fare is being skewed by large outliers. Both the cabin and ticket 
variables have many categories with low counts that make them hard to visualize 
with a bar plot. However, the `distributions()` function works great for the 
other variables as a means of quick visual analysis.

### Thoughts on `xray` package

The functions in the `xray` package provide a simple way to quickly glean some 
insight from unknown data sets. By letting the user pull out information on the 
anomalies and distributions of variables, the user is able to easily compare 
the variables and determine which ones require further investigation. It is a 
wonderful stepping off point for deeper analysis and I look forward to see the 
package develop further.
